Open-source tools in R for landscape ecology

### 2021-05-03 paper/paper.Rmd
Making changes to manuscript addressing R2

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we examine existing packages for downloading, processing and visualisation of spatial data, as well as, those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
---

@@

base_size <- 7.5
```

@@

Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is essential and ecological processes can vary spatially \cite{With2019}.
Thus, landscape ecology emphasizes \textcolor{blue}{spatial} patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these \textcolor{blue}{spatial patterns}.

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow to use, modify and distribute the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, open-source software facilitates (computational) reproducibility and allows a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratizes scientific research as free-of-cost software removes a gatekeeper for researchers without access to proprietary software.
For more general overviews of open-source software for \textcolor{blue}{landscape ecology studies} see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

One successful example of an open-source project is the \textit{R} programming language, and its extensions called packages \cite{RCoreTeam2019}. 
\textcolor{blue}{Packages allow to easily share code and make functionality and documentation available to users and by that extent the capability of the \textit{R} programming language \cite{Wickham2015}}.
Firstly released in 1995 \cite{Smith2016}, \textcolor{blue}{this} programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a large number of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
Thus, the \textit{R} programming language can be a very powerful tool. 
In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \textit{R} environment (Fig. \@ref(fig:fig-workflow)).

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review, we focus on software implemented in the \textit{R} programming language.
\textcolor{blue}{We acknowledge that there are reasons and circumstances also to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.}
Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
\textcolor{blue}{We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared.  
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.}
In the second part, we present a survey in which we asked the community how they currently use the \textit{R} programming language and to identify topics for which \textit{R} packages are presently missing for landscape ecology.

@@

Most \textit{R} packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
There are three major online platforms to host \textit{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
The last one focuses on tools for the analysis of genomic data; \textcolor{blue}{therefore, in this review we focus on only the former two.}

@@

The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}. 
At the same time, it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of \textit{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \textit{R} packages, i.e., package \textit{x} requires and uses code from package \textit{y}.
CRAN also provides archived versions of outdated or orphaned packages, thus ensuring long-term availability and reproducibility of code.
That being said, most R packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
Furthermore, online communities like \textit{rOpenSci} also provide a peer-review process for code quality.
It is important to mention that while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backward compatibility is not always ensured.
Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
\textcolor{blue}{These packages facilitate reproducibility to a high degree by preserving the project environment, including specific package versions used for the analysis.}
For more information about \textit{R} package development in general, see \cite{Wickham2015}.

@@

While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data. 
\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.} 
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.

@@

\textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for storing the CRS.
Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.}

@@

Finally, there are several tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
\textcolor{blue}{Direct access to these GIS software within \textit{R} allows to create reproducible and sharable \textit{R} scripts, even if functionality is currently not available in \textit{R}.}

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
\textit{R} allows two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user can modify the map by, e.g., changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
However, the generic functions are focused on quick visual inspection of the data (Fig. \@ref(fig:fig-map)a), rather than the creating complete maps \textcolor{blue}{as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.}

The popular plotting package \textit{ggplot2} (Fig. \@ref(fig:fig-map)b) \textcolor{blue}{can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
Additionally, there is an extension especially designed to plot spatial data named \textit{ggspatial}, which provides additional map elements \cite{Dunnington2020}.
For more information about \textit{ggplot2} in general, see \cite{Wickham2016}}.
The \textit{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
\textcolor{blue}{This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.}
Quick interactive visualization of spatial data can be done with the \textit{mapview} package\cite{Appelhans2020}.
Both, \textit{tmap} and \textit{mapview} build upon the \textit{leaflet} package and leaflet javascript library \cite{Cheng2021}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \textit{cartography} package \cite{Giraud2016}.
Further plotting methods for raster objects can be found in the \textit{rasterVis} package \cite{Lamigueiro2020}.
A slightly different approach to visualizing spatial data is adapted by the \textit{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

The introduction of the \textit{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \textit{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics in within the \textit{R} environment.

@@

Distance-based methods allow to detect and correct for spatial autocorrelation in data.
It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
The \textit{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
The \textit{rinla} \cite{Rue2009} (\url{https://www.r-inla.org}) and \textit{inlabru} \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.

@@

Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
Originated in the 1970s, SDM has experienced numerous methodological advancement, and a numerous body of literature exists today \cite{Zimmermann2010,Norberg2019}.
Additionally, textbooks introducing basic concepts of SDM in \textit{R} exists \cite{Guisan2017,Fletcher2018}.

@@

Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across \textcolor{blue}{a couple} \textit{R} packages.
A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \textit{NLMR} package \cite{Sciaini2018}.
Furthermore, the \textit{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.

@@

```{r participation, include = FALSE}
digits <- 1

# number of people that took part
n_people <- nrow(answers)

@@

To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community. \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey}.
All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).

In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
Other, less frequent answers were \textcolor{blue}{"Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).}

@@

# create data.frame
tasks_df <- tibble(tasks = important_tasks) %>%
  dplyr::group_by(tasks) %>%
  dplyr::summarise(n = dplyr::n()) %>%
  dplyr::mutate(n_rel = n / sum(n) * 100,
                tasks_reclass = dplyr::case_when(n >= 5 ~ tasks,
                                                 n < 5 ~ "others")) %>%
  dplyr::group_by(tasks_reclass) %>%
  dplyr::summarise(n_rel = sum(n_rel)) %>%
  dplyr::mutate(label = paste0(round(n_rel, digits = 1), "%"),
                col = dplyr::case_when(tasks_reclass == "others" ~ 0,
                                       tasks_reclass != "others" ~ 1),
                col = factor(col)) %>%
  dplyr::arrange(-n_rel)

@@

Next, we were interested in the most important tasks to the workflow of the participants.
\textcolor{blue}{Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options} (Fig. \@ref(fig:fig-survey)b).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").

@@

More people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
This was also represented in the most used \textit{R} packages (Fig. \@ref(fig:fig-survey)c).
When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
The \textit{raster} package was mentioned the most, followed by the \textit{sf} package.
Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey)c.
Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], 2)`%).

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we examine existing packages for downloading, processing and visualisation of spatial data, as well as those specifically developed for spatial ecological analysis. Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
---

@@

base_size <- 7.5

digits <- 1
```

@@

Landscape ecology focuses on how ecological processes are influenced by the heterogeneous landscapes and how the ecological processes themselves influence the landscapes \cite{Turner1989,Turner2005,With2019}.
Because landscapes are defined as mosaics of different land covers, ecosystems, habitat types, or land uses \cite{Forman1986,Forman1995,Wiens1995a}, spatial context is essential and ecological processes can vary spatially \cite{With2019}.
Thus, landscape ecology emphasizes spatial patterns to a high degree \cite{Risser1984} and consequently relies on software to preprocess, modify, model, analyze, and visualize these spatial patterns.

Software to manage and analyze data becomes increasingly important in modern scientific research \cite{Wilson2014} and many scientific studies would not have been possible without open-source software \cite{Prlic2012}.
Open-source software includes all software that is released under licenses that allow for the use, modification and distribution of the software \cite{St.Laurent2008}.
Open-source software development has many advantages, such as fast innovation, transparency, reliability, and longevity, mainly due to many diverse contributors \cite{vonKrogh2006,St.Laurent2008}.
Additionally, open-source software facilitates (computational) reproducibility and allows a better understanding of the methodology of a study \cite{Prlic2012,Powers2019}.
Furthermore, open-source software allows other scientists to reuse code and not "reinvent the wheel" \cite{Prlic2012} by customizing existing software to their specific needs \cite{Steiniger2009}.
Importantly, though not strictly necessary by definition of open-source \cite{Steiniger2009,Steiniger2009a}, most open-source software is also free-of-cost, in contrast to often expensive proprietary software \cite{vonKrogh2006,Steiniger2009,Steiniger2009a}.
This democratizes scientific research.
For more general overviews of open-source software for landscape ecology studies see \cite{Jolma2008,Steiniger2009,Steiniger2009a,Istvan2012}.

One successful example of an open-source project is the \textit{R} programming language, and its extensions called packages \cite{RCoreTeam2019}. 
Packages allow to easily share code and make functionality and documentation available to users and by that extend the capability of the \textit{R} programming language \cite{Wickham2015}.
First released in 1995 \cite{Smith2016}, this programming language is among the most popular programming languages today, especially in ecology \cite{Lai2019}.
Originally introduced as a statistical programming language, a large number of packages designed to analyze spatial data subsequently emerged for the \textit{R} programming language \cite{Bivand2006,Lovelace2019}.
Thus, the \textit{R} programming language can be a very powerful tool. 
In addition to handling spatial data, other analytical tasks such as statistical modeling, creation of publication-ready figures, and even complete reports can be done within the \textit{R} environment (Fig. \@ref(fig:fig-workflow)).

@@

Even though many other open-source tools \cite{QGISDevelopmentTeam2016,GRASSDevelopmentTeam2017,Porta2017} and suitable programming languages (e.g., Python) for landscape ecology exist, in this review, we focus on software implemented in the \textit{R} programming language.
We acknowledge that there are reasons and circumstances to use other software tools and programming languages, however, a comprehensive discussion of pros and cons is outside the scope of this manuscript.
Thus, in the first part of this article, we present an overview of existing \textit{R} packages for landscape ecology and closely related fields (Table \@ref(tab:packages)).
We included only packages that were available on the official \textit{R} package archive network CRAN at the time the manuscript was prepared.  
This ensures that all included packages were at least maintained to fulfill CRAN's technical quality standards.
In the second part, we present a survey in which we asked the community how they currently use the \textit{R} programming language and to identify topics for which \textit{R} packages are presently missing for landscape ecology.

@@

Most \textit{R} packages are developed and maintained by the community, which shows how open-source software development can facilitate innovation, reproducibility, and reuse of code.
There are three major online platforms to host \textit{R} packages and make them accessible to potential users: CRAN, GitHub, and Bioconductor.
The last one focuses on tools for the analysis of genomic data; therefore, in this review we focus on only the former two.

@@

The guaranteed technical quality standard of \textit{CRAN} requires more initial work for developers compared to \textit{GitHub}. 
At the same time, it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard of \textit{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \textit{R} packages, i.e., package \textit{x} requires and uses code from package \textit{y}.
CRAN also provides archived versions of outdated or orphaned packages, thus ensuring long-term availability and reproducibility of code.
That being said, most R packages can be found on both platforms, and many developers use \textit{GitHub} for regular development and \textit{CRAN} to publish stable releases of the packages.
Furthermore, online communities like \textit{rOpenSci} also provide a peer-review process for code quality.
It is important to mention that while the package environment has many advantages, its highly dynamic characteristic with constant updates by the community might also be a threat to reproducibility since backward compatibility is not always ensured.
Packages that deal with such issue include \textit{groundhog} \cite{Simonsohn2021}, \textit{packrat} \cite{Ushey2018}, or \textit{renv} \cite{Ushey2020}.
These packages facilitate reproducibility to a high degree by preserving the project environment, including specific package versions used for the analysis.
For more information about \textit{R} package development in general, see \cite{Wickham2015}.

@@

While base \textit{R} has several built-in data structures, including vectors, matrices, data frames, and lists, yet it has no internal support for reading, processing, or visualizing spatial data. 
However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with higher computational cost \cite{Lovelace2019}.
The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \textit{R} packages directly depend on or import \textit{sp}.
However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.
\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for storing the CRS.
Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.

@@

Finally, there are several tools for landscape ecology implemented in GIS software, such as `r.li` or `r.pi` for GRASS GIS \cite{Wegmann2018,Neteler2012,Porta2017}, terrain analysis methods in SAGA GIS \cite{Conrad2015}, or morphological operations for Google Earth Engine.
It is possible to control several GIS software directly from \textit{R} using dedicated packages, such as \textit{rgrass7} \cite{Bivand2021} for GRASS GIS, \textit{RSAGA} \cite{Brenning2018} for SAGA GIS, and \textit{rgee} \cite{Aybar2020} for Google Earth Engine.
Direct access to these GIS software within \textit{R} allows to create reproducible and sharable \textit{R} scripts, even if functionality is currently not available in \textit{R}.

@@

Creating maps is essential when working with spatial data.
Maps play an important role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
\textit{R} allows two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user can modify the map by, e.g., changing the displayed values.
All packages listed in Section \@ref(sec:spatrep) have build-in methods for plotting spatial data using the generic `plot()` function (Fig. \@ref(fig:fig-map)a).
However, the generic functions are focused on quick visual inspection of the data (Fig. \@ref(fig:fig-map)a), rather than the creating complete maps as they do not directly support additional map elements (e.g., scale bar, north arrow) nor small multiples maps.

The popular plotting package \textit{ggplot2} (Fig. \@ref(fig:fig-map)b) can also visualize spatial data and is build on a layered grammar of graphics, which allows to create maps by combining individual graphical elements (e.g., raster and vector elements) \cite{Wickham2016}.
Additionally, there is an extension especially designed to plot spatial data named \textit{ggspatial}, which provides additional map elements \cite{Dunnington2020}.
For more information about \textit{ggplot2} in general, see \cite{Wickham2016}.
The \textit{tmap} package provides a provides a coherent plotting system intended for drawing maps (Fig. \@ref(fig:fig-map)c) \cite{Tennekes2018}. 
This package operates in two modes, static and interactive, which means that the same code can be used to create static or interactive maps.
Quick interactive visualization of spatial data can be done with the \textit{mapview} package\cite{Appelhans2020}.
Both, \textit{tmap} and \textit{mapview} build upon the \textit{leaflet} package and leaflet javascript library \cite{Cheng2021}.
Static thematic maps, including proportional symbols, choropleth, or typology maps, can be created with the \textit{cartography} package \cite{Giraud2016}.
Further plotting methods for raster objects can be found in the \textit{rasterVis} package \cite{Lamigueiro2020}.
A slightly different approach to visualizing spatial data is adapted by the \textit{rayshader} package \cite{Morgen-Wall2020} that creates topographic 2D and 3D maps.

@@

The introduction of \textit{FRAGSTATS} in 1995 heavily facilitated the use of landscape metrics software \cite{McGarigal2012,Kupfer2012,Gustafson2019} and the \textit{landscapemetrics} package \cite{Hesselbarth2019} allows to calculate the most widely used landscape metrics within the \textit{R} environment.

@@

Distance-based methods allow for the detection and correction of spatial autocorrelation in data.
It is key to do so as landscape data are highly spatially autocorrelated, and this non-independence can affect inferences from statistical modeling. 
The \textit{spdep} package \cite{Bivand2013} has methods for quantifying multiple metrics of spatial autocorrelation and correcting these in a spatial autoregressive model.
The \textit{rinla} \cite{Rue2009} (\url{https://www.r-inla.org}) and \textit{inlabru} \cite{Bachl2019} packages also provide functionality for modeling of spatially structured data.

@@

Species distribution modeling (SDM) examines how landscape patterns (e.g., habitat suitability or resources availability) influence and determine the patterns of species' distributions, mainly to infer ecological processes and predict future species' distributions \cite{Wiersma2011}.
Originated in the 1970s, SDM has experienced numerous methodological advancement, and a large body of literature exists today \cite{Zimmermann2010,Norberg2019}.
Additionally, textbooks introducing basic concepts of SDM in \textit{R} exist \cite{Guisan2017,Fletcher2018}.

@@

Neutral landscape models are used to create structured landscapes in the absence of specific ecological and landscape processes as null models against which hypotheses including specific ecological and landscape processes can be tested statistically \cite{Gardner1987,With1997}.
Because neutral landscape models are not based on ecological and landscape processes, many different generic algorithms to create landscapes can be found across several \textit{R} packages.
A comprehensive collection of algorithms to simulate neutral landscape models specifically designed for landscape ecology can be found in the \textit{NLMR} package \cite{Sciaini2018}.
Furthermore, the \textit{RandomFields} package \cite{Schlather2015} allows to simulate Gaussian fields, which could be used as neutral landscape models.

@@

```{r participation, include = FALSE}
# number of people that took part
n_people <- nrow(answers)

@@

To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community. 
Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.
All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).

In total, the survey was answered by `r n_people` participants, of which the majority were either "PhD students" (`r current_position[[1, 2]]`%), followed by "Post-Docs" (`r current_position[[2, 2]]`%) and "Professors" (`r current_position[[3, 2]]`%).
Other, less frequent answers were "Data scientists", "None of the above", "Government employees", "Master's degree student" and "Bachelor's degree student" (in decreasing order).

@@

# create data.frame
tasks_df <- tibble(tasks = important_tasks) %>%
  dplyr::group_by(tasks) %>%
  dplyr::summarise(n = dplyr::n()) %>%
  dplyr::mutate(n_rel = n / sum(n) * 100,
                tasks_reclass = dplyr::case_when(n >= 5 ~ tasks,
                                                 n < 5 ~ "others")) %>%
  dplyr::group_by(tasks_reclass) %>%
  dplyr::summarise(n_rel = sum(n_rel)) %>%
  dplyr::mutate(label = paste0(round(n_rel, digits = digits), "%"),
                col = dplyr::case_when(tasks_reclass == "others" ~ 0,
                                       tasks_reclass != "others" ~ 1),
                col = factor(col)) %>%
  dplyr::arrange(-n_rel)

@@

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, "(pre-)processing of data", "spatial statistics", and "creating maps" were the most selected options (Fig. \@ref(fig:fig-survey)b).
The available options seemed to describe the most important task to the workflow quite well since only very few participants selected the "others" option (all options with less than five total answers were classified as "others").

@@

More people use the raster data model (`r data_model[[1, 2]]`%) in comparison to the vector data model (`r data_model[[2, 2]]`%).
This was also represented in the most used \textit{R} packages (Fig. \@ref(fig:fig-survey)c).
When asked for the three most used packages, participants of the survey listed `r length(unique(r_packages))` packages in total.
The \textit{raster} package was mentioned the most, followed by the \textit{sf} package.
Both packages are designed for basic and advanced data handling and processing of raster and vector data, respectively, representing the results of Fig. \@ref(fig:fig-survey)c.
Nevertheless, the large availability and usage of different \textit{R} packages across the community can be seen in the large "others" option (packages mentioned by less than 5 participants; `r round(r_packages_df[[1, 2]], digits)`%).
